---
title: "r01_nc_filtering_exac_full_or"
output: html_document
editor_options: 
  chunk_output_type: console
---

Started: David Gilfoyle 3 May 2018  
Updated: David Gilfoyle 16 May 2018

---

# Summary

Variant filtering of AF filtered dataset to select non-coding (Consequence) and high confidence (HC) non-coding (FATHMM_MKL_NC > 0.5 and CADD_PHRED > 15)  
Dataset: ExAC (full)  
Gene Selection: Olfactory Receptor Genes  
Input Variants: 56,122

# start_section

```{r start_section}

Sys.time()
rm(list=ls())
graphics.off()

# Install libraries
library(dplyr)
library(knitr)

```

# load_rdata_file

```{r load_rdata_file}

# Setup input folder
input_folder="/home/ec2-user/results/filtered/af_filt/exac_v1"

# Load RData file - contains dataframe containing all variants
load(paste(input_folder, "r01_exac_full_or_af_filtered.RData", sep="/"))

# Setup script input dataframe naming
input.df <- fixed_af_filt_exac_full_or.df
rm(fixed_af_filt_exac_full_or.df)

# Check input dataframe
dim(input.df)
class(input.df)
input.df[1:20,1:10]
class(input.df$Consequence)
input.df$Consequence[1:20]
class(input.df$FATHMM_MKL_NC)
input.df$FATHMM_MKL_NC[1:20]
class(input.df$CADD_PHRED)
input.df$CADD_PHRED[1:20]
class(input.df$ExACpLI)
input.df$ExACpLI[1:20]

# Setup base and output folders
base_folder="/home/ec2-user/scripts/filtration/nc_filt/exac_v1"
output_folder="/home/ec2-user/results/filtered/nc_filt/exac_v1"

getwd()
setwd(base_folder)
getwd()

opts_knit$set(root.dir = base_folder)

```

# variants_associated_with_or_genes

```{r variants_associated_with_or_genes}

# Set of OR genes supporting variant filtration (with OR8U9 and OR9G9 removed)
or_genes <- c("OR1A1", "OR1A2", "OR1C1", "OR1D2", "OR1D5", "OR1E1", "OR1E2", "OR1F1", "OR1F12", "OR1G1", "OR1I1", "OR1J1", "OR1J2", "OR1J4", "OR1K1", "OR1L1", "OR1L3", "OR1L4", "OR1L6", "OR1L8", "OR1M1", "OR1N1", "OR1N2", "OR1Q1", "OR1S2", "OR2A1", "OR2A12", "OR2A14", "OR2A2", "OR2A25", "OR2A4", "OR2A42", "OR2A5", "OR2A7", "OR2AE1", "OR2AG2", "OR2AJ1", "OR2AK2", "OR2AP1", "OR2AT4", "OR2B11", "OR2B2", "OR2B3", "OR2B6", "OR2C1", "OR2C3", "OR2D2", "OR2D3", "OR2F2", "OR2G2", "OR2G3", "OR2G6", "OR2H1", "OR2H2", "OR2J2", "OR2J3", "OR2K2", "OR2L13", "OR2L2", "OR2L3", "OR2L5", "OR2M2", "OR2M3", "OR2M4", "OR2M5", "OR2M7", "OR2T1", "OR2T10", "OR2T12", "OR2T2", "OR2T27", "OR2T29", "OR2T3", "OR2T33", "OR2T34", "OR2T35", "OR2T4", "OR2T5", "OR2T6", "OR2T7", "OR2T8", "OR2V1", "OR2V2", "OR2W1", "OR2W3", "OR2Y1", "OR2Z1", "OR3A1", "OR3A2", "OR3A3", "OR4A15", "OR4A16", "OR4A47", "OR4A5", "OR4B1", "OR4C11", "OR4C12", "OR4C13", "OR4C15", "OR4C3", "OR4C46", "OR4C6", "OR4D1", "OR4D10", "OR4D11", "OR4D2", "OR4D5", "OR4D6", "OR4D9", "OR4E2", "OR4F15", "OR4F16", "OR4F17", "OR4F21", "OR4F29", "OR4F3", "OR4F4", "OR4F5", "OR4F6", "OR4K1", "OR4K13", "OR4K14", "OR4K15", "OR4K17", "OR4K2", "OR4K5", "OR4L1", "OR4M1", "OR4M2", "OR4N2", "OR4N4", "OR4N5", "OR4P4", "OR4Q3", "OR4S1", "OR4S2", "OR5A1", "OR5A2", "OR5AC2", "OR5AK2", "OR5AN1", "OR5AP2", "OR5AS1", "OR5AU1", "OR5B12", "OR5B17", "OR5B2", "OR5B21", "OR5B3", "OR5C1", "OR5D14", "OR5D16", "OR5D18", "OR5F1", "OR5H1", "OR5H14", "OR5H15", "OR5H2", "OR5I1", "OR5J2", "OR5K1", "OR5K2", "OR5K3", "OR5K4", "OR5L2", "OR5M1", "OR5M10", "OR5M11", "OR5M3", "OR5M8", "OR5M9", "OR5P2", "OR5P3", "OR5T1", "OR5T2", "OR5T3", "OR5V1", "OR5W2", "OR6A2", "OR6B1", "OR6B2", "OR6B3", "OR6C1", "OR6C2", "OR6C3", "OR6C4", "OR6C6", "OR6C65", "OR6C68", "OR6C70", "OR6C74", "OR6C75", "OR6C76", "OR6F1", "OR6K2", "OR6K3", "OR6K6", "OR6M1", "OR6N1", "OR6N2", "OR6P1", "OR6S1", "OR6T1", "OR6V1", "OR6X1", "OR6Y1", "OR7A10", "OR7A17", "OR7A5", "OR7C1", "OR7C2", "OR7D2", "OR7D4", "OR7E24", "OR7G1", "OR7G2", "OR7G3", "OR8A1", "OR8B12", "OR8B2", "OR8B3", "OR8B8", "OR8D1", "OR8D4", "OR8G5", "OR8H1", "OR8H2", "OR8H3", "OR8I2", "OR8J1", "OR8J3", "OR8K1", "OR8K5", "OR8S1", "OR8U1", "OR8U8", "OR9A2", "OR9A4", "OR9G1", "OR9G4", "OR9I1", "OR9K2", "OR9Q1", "OR9Q2", "OR10A2", "OR10A3", "OR10A4", "OR10A5", "OR10A7", "OR10AD1", "OR10AG1", "OR10G2", "OR10G3", "OR10G4", "OR10G6", "OR10G7", "OR10G8", "OR10G9", "OR10H1", "OR10H2", "OR10H3", "OR10H4", "OR10H5", "OR10J1", "OR10J3", "OR10J5", "OR10K1", "OR10K2", "OR10P1", "OR10Q1", "OR10R2", "OR10S1", "OR10T2", "OR10V1", "OR10W1", "OR10Z1", "OR11A1", "OR11G2", "OR11H1", "OR11H12", "OR11H2", "OR11H4", "OR11H6", "OR11L1", "OR12D3", "OR13A1", "OR13C2", "OR13C3", "OR13C4", "OR13C5", "OR13C8", "OR13C9", "OR13D1", "OR13F1", "OR13G1", "OR13H1", "OR13J1", "OR14A16", "OR14A2", "OR14C36", "OR14I1", "OR14J1", "OR14K1", "OR51A2", "OR51A4", "OR51A7", "OR51B4", "OR51B5", "OR51B6", "OR51D1", "OR51E1", "OR51E2", "OR51F2", "OR51G2", "OR51H1", "OR51I1", "OR51I2", "OR51L1", "OR51M1", "OR51S1", "OR51T1", "OR51V1", "OR52A1", "OR52A5", "OR52B2", "OR52B6", "OR52D1", "OR52E2", "OR52E4", "OR52E5", "OR52E6", "OR52E8", "OR52H1", "OR52I1", "OR52I2", "OR52J3", "OR52K1", "OR52K2", "OR52L1", "OR52M1", "OR52N1", "OR52N2", "OR52N5", "OR52W1", "OR56A1", "OR56A3", "OR56A4", "OR56A5", "OR56B1", "OR56B4")

# Filtering of variants based on DDR genes using SYMBOL 
output.df <- input.df %>%
  filter(input.df$SYMBOL %in% or_genes | input.df$NEAREST %in% or_genes)

removed_variant.df <- input.df %>%
  filter(!input.df$SYMBOL %in% or_genes | !input.df$NEAREST %in% or_genes)

# Check output dataframe
dim(output.df)
class(output.df)
output.df[1:20,1:10]
class(output.df$SYMBOL)
unique(output.df$SYMBOL)
unique(removed_variant.df$SYMBOL)

```

# nc_filtering

Filtering based on VEP consequences:  
1) 5_prime_UTR_variant  
2) 3_prime_UTR_variant  
3) upstream_gene_variant  
4) downstream_gene_variant  
5) TF_binding_site_variant  
6) regulatory_region_variat  

```{r nc_filtering}

# Filtration based just on consequence
output.df <- output.df %>% 
  filter(grepl("5_prime_UTR_variant",Consequence) |
           grepl("3_prime_UTR_variant",Consequence) |
           grepl("upstream_gene_variant",Consequence) |
           grepl("downstream_gene_variant",Consequence) |
           grepl("TF_binding_site_variant",Consequence) |
           grepl("regulatory_region_variant",Consequence))

# Check output dataframe
dim(output.df)
output.df[1:20,1:10]
output.df$Consequence[1:20]
output.df$FATHMM_MKL_NC[1:20]
output.df$CADD_PHRED[1:20]

```

# additional_filtering

Output assessed based on further filtering with:  
1) FATHMM_MKL_NC > 0.5 - assessment indicates that some entries are NA - these appear to be insertion, deletion and substitution variants and they will be retained. However, removal of variants with FATHMM_MKL_NC > 0.5 is considered appropriate to enrich dataset for potentially deleterious variants  
2) CADD_PHRED > 20 - assessment indicates that some entries are NA - these appear to be mainly insertion, deletion and substitution variants and they will be retained. However, removal of variants with CADD_PHRED <= 20 is considered appropriate to enrich dataset for potentially deleterious variants


```{r additional_filtering}

# Assessment of filtering FATHMM_MKL_NC > 0.5
class(output.df$FATHMM_MKL_NC)
sum(is.na(output.df$FATHMM_MKL_NC))
sum(as.numeric(output.df$FATHMM_MKL_NC) == 0.5, na.rm = TRUE)
sum(as.numeric(output.df$FATHMM_MKL_NC) < 0.5, na.rm = TRUE)
sum(as.numeric(output.df$FATHMM_MKL_NC) > 0.5, na.rm = TRUE)

# output_fathmm_na.df <- output.df %>% 
#   filter(is.na(FATHMM_MKL_NC))

# Assessment of filtering CADD_PHREDD > 20
sum(is.na(output.df$CADD_PHRED))
sum(as.numeric(output.df$CADD_PHRED) == 20, na.rm = TRUE)
sum(as.numeric(output.df$CADD_PHRED) < 20, na.rm = TRUE)
sum(as.numeric(output.df$CADD_PHRED) > 20, na.rm = TRUE)

# output_cadd_na.df <- output.df %>% 
#  filter(is.na(CADD_PHRED))

# Additional filtration based on FATHMM_MKL_NC and CADD_PHRED
output_hc.df <- output.df %>% 
  filter((as.numeric(FATHMM_MKL_NC) > 0.5) |
           (is.na(FATHMM_MKL_NC))) %>%
  filter((as.numeric(CADD_PHRED) > 20) |
           (is.na(CADD_PHRED)))

# Check output dataframe
dim(output_hc.df)
output_hc.df[1:20,1:10]
output_hc.df$Consequence[1:20]
output_hc.df$FATHMM_MKL_NC[1:20]
output_hc.df$CADD_PHRED[1:20]
unique(output_hc.df$SYMBOL)

```

# update_dataframe_naming

```{r update_dataframe_naming}

# Reset of output dataframe names
fixed_nc_filt_exac_full_or.df <- output.df
fixed_nc_hc_filt_exac_full_or.df <- output_hc.df

# Clean-up
rm(input.df, output.df, output_hc.df, or_genes, removed_variant.df)

```

# save_results

```{r save_results}

save.image(paste(output_folder, "r01_exac_full_or_nc_filtered.RData", sep="/"))

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```

