---
title: "r01_variant_filt_analysis_kgen_or"
output: html_document
editor_options: 
  chunk_output_type: console
---

Started: David Gilfoyle 2 May 2018  
Updated: David Gilfoyle 8 June 2018

---

# Summary

Data Analysis Based on AC Filtered Dataset 
Dataset: ICR1000  
Gene Selection: Olfactory Receptor Genes  
Input Variants: 59,262

# start_section

```{r start_section}

Sys.time()
rm(list=ls())
graphics.off()

# Install libraries
library(dplyr)
library(ggplot2)
library(knitr)

```

# load_rdata_file

```{r load_rdata_file}

# Setup input folder
input_folder="/home/ec2-user/results/filtered/ac_filt/kgen"

# Load RData file - contains dataframe containing all variants
load(paste(input_folder, "r01_kgen_or_ac_filtered.RData", sep="/"))

# Set data frame variable
fixed.df <- fixed_ac_filt_kgen_or.df

# Remove imported dataset
rm (fixed_ac_filt_kgen_or.df, gt_ac_filt_kgen_or.mx)

# Check input dataframe
dim(fixed.df)
class(fixed.df)
str(fixed.df)
fixed.df[1:20,1:10]

# Setup base and output folders
base_folder="/home/ec2-user/scripts/data_analysis/variant_analysis/kgen/or"
output_folder="/home/ec2-user/results/data_analysis/variant_analysis/kgen/or"

getwd()
setwd(base_folder)
getwd()

opts_knit$set(root.dir = base_folder)

```

# variant_class_analysis

```{r variant_class_analysis}

# Check for missed data
sum(is.na(fixed.df$VARIANT_CLASS)) # should be 0 (no missed data)

# Isoltate variant class column
varc_annotations.vector <- fixed.df$VARIANT_CLASS

# Determine different variant classes present
unique(varc_annotations.vector)

# Variant class counts
varc_counts.named.vector <- sort(table(varc_annotations.vector), decreasing = T)
varc_counts.named.vector

# make a dataframe for plotting
varc_counts.df <- data.frame(
  variant_type=names(varc_counts.named.vector),
  variant_count=as.numeric(varc_counts.named.vector))
varc_counts.df

# Compile file name
file_name <- paste(output_folder,"variant_class_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(varc_counts.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Order of levels defines the order of bars on the plot
levels(varc_counts.df$variant_type)
varc_counts.df$variant_type <- factor(varc_counts.df$variant_type, 
  levels=c("SNV", "insertion", "deletion", "indel", "substitution"))
levels(varc_counts.df$variant_type)

# Pie chart of variant class data
pie <- ggplot(varc_counts.df, aes(x = "", y = variant_count, fill = factor(variant_type))) +
  geom_bar(width = 1, stat = "identity") +
  theme(axis.line = element_blank(), plot.title = element_text(hjust=0.5)) +
  labs(fill = "Variant Class", x = NULL, y = NULL) +
  ggtitle("1000 Genomes Project - Number of Variants by Class")
pie + coord_polar(theta = "y", start = 0)

ggsave("variant_class_counts_plot.pdf", output_folder, plot = last_plot(), device = "pdf", width = 297, height = 210, units = "mm", dpi = 300)

# Clean-up
rm(pie, varc_counts.df, file_name, varc_annotations.vector, varc_counts.named.vector)

```

# consequence_analysis

```{r consequence_analysis}

# Check for missed data (is.na produces a logical vector)
sum(is.na(fixed.df$Consequence)) # should be 0 (no missed data)

# Split multiple consequences
csq_annotations.list <- strsplit(as.vector(fixed.df$Consequence),"&")

# Flatten list of string vectors to a string vector
csq_annotations.vector <- as.vector(unlist(csq_annotations.list))

# What different consequences were observed?
unique(csq_annotations.vector)

# Variant class counts
csq_counts.named.vector <- sort(table(csq_annotations.vector), decreasing = T)
csq_counts.named.vector

# make a dataframe for plotting
csq_counts.df <- data.frame(
  csq_type=names(csq_counts.named.vector),
  csq_count=as.numeric(csq_counts.named.vector))
csq_counts.df

# Compile file name
file_name <- paste(output_folder,"consequence_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(csq_counts.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Order of levels defines the order of bars on the plot
levels(csq_counts.df$csq_type)
csq_counts.df$csq_type <- factor(csq_counts.df$csq_type, 
  levels=as.vector(csq_counts.df$csq_type))
levels(csq_counts.df$csq_type)

# Bar graph of consequence data
ggplot(csq_counts.df, aes(csq_type, csq_count)) +
       geom_bar(stat="identity", fill="blue", alpha=0.5) + 
       geom_text(aes(label=format(csq_count, big.mark=","), vjust=-1)) +
       theme(axis.text.x = element_text(hjust=1, size=12, angle=75)) +
       ylim(0,18000) + 
       labs(x = "", y = "Number of Variants") + 
       ggtitle("1000 Genomes Project - Number of Variants by Consequence")

ggsave("consequence_counts_plot.pdf", output_folder, plot = last_plot(), device = "pdf", width = 297, height = 210, units = "mm", dpi = 300)

# Clean-up
rm(csq_counts.df, file_name, csq_annotations.list, csq_annotations.vector, csq_counts.named.vector)

```

# sift_consequence_analysis

```{r sift_consequence_analysis}

# Count number of NA values
sum(is.na(fixed.df$SIFT))

# Isolate SIFT type and score column
sift_annotations.df <- data.frame(SIFT=c(fixed.df$SIFT))

# Seperate SIFT type and score into two seperare columns
sift_annotations.df <- 
  sift_annotations.df %>% 
  mutate(sift_type=sub("\\(.*\\)","",SIFT)) %>% 
  mutate(sift_score=as.numeric(
    sub(".*\\(","", sub("\\)","",SIFT)))) %>% 
  select(-SIFT)

# SIFT consequence counts
sift_counts <- sort(table(sift_annotations.df$sift_type), decreasing = T)

# Make a dataframe for table
sift_type_counts.df <- data.frame(
  sift_type=names(sift_counts),
  sift_count=as.numeric(sift_counts))
sift_type_counts.df

# Compile file name
file_name <- paste(output_folder,"sift_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(sift_type_counts.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Clean-up
rm(sift_annotations.df, sift_type_counts.df, sift_counts, file_name)

```

# polyphen_consequence_analysis

```{r polyphen_consequence_analysis}

# Count number of NA values
sum(is.na(fixed.df$PolyPhen))

# Isolate PolyPhen type and score column
poly_annotations.df <- data.frame(poly=c(fixed.df$PolyPhen))

# Seperate PolyPhen type and score into two seperare columns
poly_annotations.df <- 
  poly_annotations.df %>% 
  mutate(polyphen_type=sub("\\(.*\\)","",poly)) %>% 
  mutate(polyphen_score=as.numeric(
    sub(".*\\(","", sub("\\)","",poly)))) %>% 
  select(-poly)

# PolyPhen consequence counts
poly_counts <- sort(table(poly_annotations.df$polyphen_type), decreasing = T)

# Make a dataframe for table
poly_type_counts.df <- data.frame(
  polyphen_type=names(poly_counts),
  polyphen_count=as.numeric(poly_counts))
poly_type_counts.df

# Compile file name
file_name <- paste(output_folder,"polyphen_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(poly_type_counts.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Clean-up
rm(poly_annotations.df, poly_type_counts.df, poly_counts, file_name)

```

# allele_frequency_distribution_analysis

```{r allele_frequency_distribution_analysis}

# Check for missed data (is.na produces a logical vector)
sum(is.na(fixed.df$AF)) # should be 0 (no missed data)

# Isolate allele frequency column
af.vector <- as.numeric(fixed.df$AF)

# Sum of extreamly rare variants with AF <= 0.0001
af_ex_rare <- sum(af.vector <= 0.0001)

# Sum of very rare variants with AF <= 0.001
af_v_rare <- sum(af.vector <= 0.001)

# Sum of rare variants with AF <= 0.01
af_rare <- sum(af.vector <= 0.01)

# Sum of low frequency variants with AF > 0.01 and < 0.05
af_low <- sum(af.vector > 0.01 & af.vector < 0.05)

# Sum of common variants with AF >= 0.05
af_common <- sum(af.vector >= 0.05)

# Prepare AF dataframe
af_dist.df <- data.frame(af_range=c("AF >= 0.05","AF > 0.01 & < 0.05","AF <= 0.01","AF <= 0.001","AF <= 0.0001"),
                         af_count=c(af_common, af_low, af_rare, af_v_rare, af_ex_rare))
af_dist.df

# Compile file name
file_name <- paste(output_folder,"af_range_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(af_dist.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Clean-up
rm(af.vector, af_dist.df, af_common, af_low, af_rare, af_v_rare, af_ex_rare, file_name)

```

# singletons

```{r singletons}

# Check for missed data (is.na produces a logical vector)
sum(is.na(fixed.df$AC)) # should be 0 (no missed data)

# Isolate dataset singleton allele count column
ac.vector <- as.numeric(fixed.df$AC)

# Sum of allele counts = 1 (singletons)
ac_sing <- sum(ac.vector == 1)

# Prepare AC dataframe
ac_sing.df <- data.frame(af_type=c("Singletons"),
                         af_count=c(ac_sing))
ac_sing.df

# Compile file name
file_name <- paste(output_folder,"ac_sing_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(ac_sing.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Clean-up
rm(ac_sing.df, ac_sing, ac.vector, file_name)

```

# variants_by_all_genes

```{r variants_by_all_genes}

# Isolate SYMBOL column
gene_variants.df <- data.frame(SYMBOL=c(fixed.df$SYMBOL))

# Calculate numbers of variants associated with each gene
variant_counts <- sort(table(gene_variants.df$SYMBOL), decreasing = T)

# Make a dataframe for table
gene_variant_counts.df <- data.frame(
  gene=names(variant_counts),
  variant_count=as.numeric(variant_counts))
gene_variant_counts.df

# Compile file name
file_name <- paste(output_folder,"all_gene_variant_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(gene_variant_counts.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Clean-up
rm(gene_variants.df, variant_counts, gene_variant_counts.df, file_name)

```

# variants_associated_with_or_genes

```{r variants_associated_with_or_genes}

# Set of OR genes supporting variant filtration (with OR8U9 and OR9G9 removed)
or_genes <- c("OR1A1", "OR1A2", "OR1C1", "OR1D2", "OR1D5", "OR1E1", "OR1E2", "OR1F1", "OR1F12", "OR1G1", "OR1I1", "OR1J1", "OR1J2", "OR1J4", "OR1K1", "OR1L1", "OR1L3", "OR1L4", "OR1L6", "OR1L8", "OR1M1", "OR1N1", "OR1N2", "OR1Q1", "OR1S2", "OR2A1", "OR2A12", "OR2A14", "OR2A2", "OR2A25", "OR2A4", "OR2A42", "OR2A5", "OR2A7", "OR2AE1", "OR2AG2", "OR2AJ1", "OR2AK2", "OR2AP1", "OR2AT4", "OR2B11", "OR2B2", "OR2B3", "OR2B6", "OR2C1", "OR2C3", "OR2D2", "OR2D3", "OR2F2", "OR2G2", "OR2G3", "OR2G6", "OR2H1", "OR2H2", "OR2J2", "OR2J3", "OR2K2", "OR2L13", "OR2L2", "OR2L3", "OR2L5", "OR2M2", "OR2M3", "OR2M4", "OR2M5", "OR2M7", "OR2T1", "OR2T10", "OR2T12", "OR2T2", "OR2T27", "OR2T29", "OR2T3", "OR2T33", "OR2T34", "OR2T35", "OR2T4", "OR2T5", "OR2T6", "OR2T7", "OR2T8", "OR2V1", "OR2V2", "OR2W1", "OR2W3", "OR2Y1", "OR2Z1", "OR3A1", "OR3A2", "OR3A3", "OR4A15", "OR4A16", "OR4A47", "OR4A5", "OR4B1", "OR4C11", "OR4C12", "OR4C13", "OR4C15", "OR4C3", "OR4C46", "OR4C6", "OR4D1", "OR4D10", "OR4D11", "OR4D2", "OR4D5", "OR4D6", "OR4D9", "OR4E2", "OR4F15", "OR4F16", "OR4F17", "OR4F21", "OR4F29", "OR4F3", "OR4F4", "OR4F5", "OR4F6", "OR4K1", "OR4K13", "OR4K14", "OR4K15", "OR4K17", "OR4K2", "OR4K5", "OR4L1", "OR4M1", "OR4M2", "OR4N2", "OR4N4", "OR4N5", "OR4P4", "OR4Q3", "OR4S1", "OR4S2", "OR5A1", "OR5A2", "OR5AC2", "OR5AK2", "OR5AN1", "OR5AP2", "OR5AS1", "OR5AU1", "OR5B12", "OR5B17", "OR5B2", "OR5B21", "OR5B3", "OR5C1", "OR5D14", "OR5D16", "OR5D18", "OR5F1", "OR5H1", "OR5H14", "OR5H15", "OR5H2", "OR5I1", "OR5J2", "OR5K1", "OR5K2", "OR5K3", "OR5K4", "OR5L2", "OR5M1", "OR5M10", "OR5M11", "OR5M3", "OR5M8", "OR5M9", "OR5P2", "OR5P3", "OR5T1", "OR5T2", "OR5T3", "OR5V1", "OR5W2", "OR6A2", "OR6B1", "OR6B2", "OR6B3", "OR6C1", "OR6C2", "OR6C3", "OR6C4", "OR6C6", "OR6C65", "OR6C68", "OR6C70", "OR6C74", "OR6C75", "OR6C76", "OR6F1", "OR6K2", "OR6K3", "OR6K6", "OR6M1", "OR6N1", "OR6N2", "OR6P1", "OR6S1", "OR6T1", "OR6V1", "OR6X1", "OR6Y1", "OR7A10", "OR7A17", "OR7A5", "OR7C1", "OR7C2", "OR7D2", "OR7D4", "OR7E24", "OR7G1", "OR7G2", "OR7G3", "OR8A1", "OR8B12", "OR8B2", "OR8B3", "OR8B8", "OR8D1", "OR8D4", "OR8G5", "OR8H1", "OR8H2", "OR8H3", "OR8I2", "OR8J1", "OR8J3", "OR8K1", "OR8K5", "OR8S1", "OR8U1", "OR8U8", "OR9A2", "OR9A4", "OR9G1", "OR9G4", "OR9I1", "OR9K2", "OR9Q1", "OR9Q2", "OR10A2", "OR10A3", "OR10A4", "OR10A5", "OR10A7", "OR10AD1", "OR10AG1", "OR10G2", "OR10G3", "OR10G4", "OR10G6", "OR10G7", "OR10G8", "OR10G9", "OR10H1", "OR10H2", "OR10H3", "OR10H4", "OR10H5", "OR10J1", "OR10J3", "OR10J5", "OR10K1", "OR10K2", "OR10P1", "OR10Q1", "OR10R2", "OR10S1", "OR10T2", "OR10V1", "OR10W1", "OR10Z1", "OR11A1", "OR11G2", "OR11H1", "OR11H12", "OR11H2", "OR11H4", "OR11H6", "OR11L1", "OR12D3", "OR13A1", "OR13C2", "OR13C3", "OR13C4", "OR13C5", "OR13C8", "OR13C9", "OR13D1", "OR13F1", "OR13G1", "OR13H1", "OR13J1", "OR14A16", "OR14A2", "OR14C36", "OR14I1", "OR14J1", "OR14K1", "OR51A2", "OR51A4", "OR51A7", "OR51B4", "OR51B5", "OR51B6", "OR51D1", "OR51E1", "OR51E2", "OR51F2", "OR51G2", "OR51H1", "OR51I1", "OR51I2", "OR51L1", "OR51M1", "OR51S1", "OR51T1", "OR51V1", "OR52A1", "OR52A5", "OR52B2", "OR52B6", "OR52D1", "OR52E2", "OR52E4", "OR52E5", "OR52E6", "OR52E8", "OR52H1", "OR52I1", "OR52I2", "OR52J3", "OR52K1", "OR52K2", "OR52L1", "OR52M1", "OR52N1", "OR52N2", "OR52N5", "OR52W1", "OR56A1", "OR56A3", "OR56A4", "OR56A5", "OR56B1", "OR56B4")

# Filtering of variants based on OR genes using SYMBOL 
or_fixed.df <- fixed.df %>%
  filter(fixed.df$SYMBOL %in% or_genes)

removed_variant.df <- fixed.df %>%
  filter(!fixed.df$SYMBOL %in% or_genes)

# Check output dataframe
dim(or_fixed.df)
class(or_fixed.df)
or_fixed.df[1:20,1:10]
class(or_fixed.df$SYMBOL)
unique(or_fixed.df$SYMBOL)
unique(removed_variant.df$SYMBOL)

```

# variants_by_or_genes

```{r variants_by_or_genes}

# Isolate SYMBOL column
gene_variants.df <- data.frame(SYMBOL=c(or_fixed.df$SYMBOL))

# Calculate numbers of variants associated with each gene
variant_counts <- sort(table(gene_variants.df$SYMBOL), decreasing = T)

# Make a dataframe for table
gene_variant_counts.df <- data.frame(
  gene=names(variant_counts),
  variant_count=as.numeric(variant_counts))
gene_variant_counts.df

# Compile file name
file_name <- paste(output_folder,"or_gene_variant_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(gene_variant_counts.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Clean-up
rm(or_genes, or_fixed.df, removed_variant.df, gene_variants.df, variant_counts, gene_variant_counts.df, file_name)

```

# dbSNP_variants

```{r dbSNP_variants}

# Count number of NA values
sum(is.na(fixed.df$ID))

# Count number of dbSNP IDs
sum(!is.na(fixed.df$ID))
sum(!fixed.df$Uploaded_variation == ".")
sum(!is.na(fixed.df$Existing_variation))

```

# clinical_significance

```{r clinical_significance}

# Count number of NA values
sum(is.na(fixed.df$CLIN_SIG))

# Split multiple consequences
clinsig_annotations.list <- strsplit(as.vector(fixed.df$CLIN_SIG),"&")
clinsig_annotations.vector <- as.vector(unlist(clinsig_annotations.list))

# Consequences summaries
clin_sig.named.vector <- unique(clinsig_annotations.vector)

# Clinical significance consequence counts
clin_sig_counts <- sort(table(clinsig_annotations.vector), decreasing = T)

# Make a dataframe for table
clin_sig_counts.df <- data.frame(
  clin_sig=names(clin_sig_counts),
  clin_sig_count=as.numeric(clin_sig_counts))
clin_sig_counts.df

# Compile file name
file_name <- paste(output_folder,"clin_sig_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(clin_sig_counts.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Clean-up
rm(clin_sig_counts.df, clinsig_annotations.list, clin_sig_counts, clin_sig.named.vector, clinsig_annotations.vector, file_name)

```

# variants_by_chromosome

```{r variants_by_chromosome}

# Check for missed data (is.na produces a logical vector)
sum(is.na(fixed.df$CHROM)) # should be 0 (no missed data)

# Isolate chromosome column
chromo.vector <- fixed.df$CHROM

# Sum of Chromosome variants
chromo_1 <- sum(chromo.vector == 1)
chromo_2 <- sum(chromo.vector == 2)
chromo_3 <- sum(chromo.vector == 3)
chromo_4 <- sum(chromo.vector == 4)
chromo_5 <- sum(chromo.vector == 5)
chromo_6 <- sum(chromo.vector == 6)
chromo_7 <- sum(chromo.vector == 7)
chromo_8 <- sum(chromo.vector == 8)
chromo_9 <- sum(chromo.vector == 9)
chromo_10 <- sum(chromo.vector == 10)
chromo_11 <- sum(chromo.vector == 11)
chromo_12 <- sum(chromo.vector == 12)
chromo_13 <- sum(chromo.vector == 13)
chromo_14 <- sum(chromo.vector == 14)
chromo_15 <- sum(chromo.vector == 15)
chromo_16 <- sum(chromo.vector == 16)
chromo_17 <- sum(chromo.vector == 17)
chromo_18 <- sum(chromo.vector == 18)
chromo_19 <- sum(chromo.vector == 19)
chromo_20 <- sum(chromo.vector == 20)
chromo_21 <- sum(chromo.vector == 21)
chromo_22 <- sum(chromo.vector == 22)
chromo_X <- sum(chromo.vector == "X")
chromo_Y <- sum(chromo.vector == "Y")

# Prepare AF dataframe
chromo_dist.df <- data.frame(chrom_number=c("1","2","3","4","5","6","7","8","9","10",
                                                 "11","12","13","14","15","16","17","18","19",
                                                 "20","21","22","X","Y"),
                         var_count=c(chromo_1, chromo_2, chromo_3, chromo_4, chromo_5, chromo_6, 
                                    chromo_7, chromo_8, chromo_9, chromo_10, chromo_11, chromo_12,
                                    chromo_13, chromo_14, chromo_15, chromo_16, chromo_17, chromo_18,
                                    chromo_19, chromo_20, chromo_21, chromo_22, chromo_X, chromo_Y))
chromo_dist.df

# Compile file name
file_name <- paste(output_folder,"chromo_counts_table.txt", sep="/")

# Write tab-separated text file
write.table(chromo_dist.df, file=file_name, quote=F, sep="\t", row.names = F, col.names = TRUE)

# Order of levels defines the order of bars on the plot
levels(chromo_dist.df$chrom_number)
chromo_dist.df$chrom_number <- factor(chromo_dist.df$chrom_number, 
  levels=c("1","2","3","4","5","6","7","8","9","10",
           "11","12","13","14","15","16","17","18","19",
           "20","21","22","X","Y"))
levels(chromo_dist.df$chrom_number)

# Bar graph of variants by chromosome data
ggplot(chromo_dist.df, aes(chrom_number, var_count)) +
       geom_bar(stat="identity", fill="blue", alpha=0.5) + 
       geom_text(aes(label=format(var_count, big.mark=","), vjust=-1)) +
       theme(axis.text.x = element_text(hjust=0.5, size=12, angle=0)) +
       ylim(0,22600) + 
       labs(x = "Chromosome", y = "Number of Variants") + 
       ggtitle("1000 Genomes Project - Number of Variants by Chromosome")

ggsave("chromo_counts_plot.pdf", output_folder, plot = last_plot(), device = "pdf", width = 297, height = 210, units = "mm", dpi = 300)

# Clean-up
rm(chromo_dist.df, chromo.vector, chromo_1, chromo_2, chromo_3, chromo_4, chromo_5, chromo_6, chromo_7,
   chromo_8, chromo_9, chromo_10, chromo_11, chromo_12, chromo_13, chromo_14, chromo_15, chromo_16, chromo_17,
   chromo_18, chromo_19, chromo_20, chromo_21, chromo_22, chromo_X, chromo_Y, file_name)

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```

