---
title: "r01_transcript_variant_filtering_ddr"
output: html_document
editor_options: 
  chunk_output_type: console
---

Started: David Gilfoyle 7 May 2018  
Updated: David Gilfoyle 8 May 2018

---

# Summary

Filtering of AF filtered dataset to select protein-coding and UTR variants (by Consequence) encomapassing gene transcripts (covering exons and UTR regions) to support assessment of normalised variants against total DDR gene transcript length  
Dataset: All  
Gene Selection: DNA Repair Genes  

# start_section

```{r start_section}

Sys.time()
rm(list=ls())
graphics.off()

# Install libraries
library(dplyr)
library(knitr)

```

# load_rdata_files

```{r load_rdata_files}

# Setup input folders
exac_input_folder="/home/ec2-user/results/filtered/af_filt/exac_v1"
icr1000_input_folder="/home/ec2-user/results/filtered/af_filt/icr1000"
kgen_input_folder="/home/ec2-user/results/filtered/af_filt/kgen"

# Load RData files - contain dataframes containing all variants (genotype matrices are removed)
load(paste(exac_input_folder, "r01_exac_full_ddr_af_filtered.RData", sep="/"))
load(paste(exac_input_folder, "r01_exac_nontcga_ddr_af_filtered.RData", sep="/"))
load(paste(icr1000_input_folder, "r01_icr1000_ddr_af_filtered.RData", sep="/"))
rm(gt_af_filt_icr1000_ddr.mx)
load(paste(kgen_input_folder, "r01_kgen_ddr_af_filtered.RData", sep="/"))
rm(gt_af_filt_kgen_ddr.mx)

# Check input dataframe
dim(fixed_af_filt_exac_full_ddr.df) # should be 99,236
dim(fixed_af_filt_exac_nontcga_ddr.df) # should be 92,077
dim(fixed_af_filt_icr1000_ddr.df) # should be 6,442
dim(fixed_af_filt_kgen_ddr.df) # should be 313,889

# Setup base folder
base_folder="/home/ec2-user/scripts/data_analysis/norm_variant"

getwd()
setwd(base_folder)
getwd()

opts_knit$set(root.dir = base_folder)

```

# variants_associated_with_ddr_genes

```{r variants_associated_with_ddr_genes}

# Set of DDR genes supporting variant filtration
ddr_genes <- c("UNG", "SMUG1", "MBD4", "TDG", "OGG1", "MUTYH", "NTHL1", "MPG", "NEIL1", "NEIL2", "NEIL3", "HMGB1", "APEX1", "APEX2", "LIG3", "XRCC1", "PNKP", "APLF", "PARP1", "PARP2", "PARP3", "MGMT", "ALKBH2", "ALKBH3", "TDP1", "TDP2", "MSH2", "MSH3", "MSH6", "MLH1", "PMS2", "MSH4", "MSH5", "MLH3", "PMS1", "PMS2P3", "XPC", "RAD23B", "CETN2", "RAD23A", "XPA", "DDB1", "DDB2", "RPA1", "RPA2", "RPA3", "ERCC3", "ERCC2", "GTF2H1", "GTF2H2", "GTF2H3", "GTF2H4", "GTF2H5", "CDK7", "CCNH", "MNAT1", "ERCC5", "ERCC1", "ERCC4", "LIG1", "RBX1", "CUL4A", "CUL4B", "ERCC8", "ERCC6", "UVSSA", "XAB2", "MMS19", "RAD51", "RAD51B", "RAD51D", "DMC1", "XRCC2", "XRCC3", "RAD52", "RAD54L", "RAD54B", "BRCA1", "SHFM1", "RAD50", "MRE11A", "NBN", "RBBP8", "MUS81", "EME1", "EME2", "SLX1A", "SLX1B", "GEN1", "BARD1", "BABAM1", "FAM175A", "BRE", "UIMC1", "BRCC3", "SYCP3", "FANCA", "FANCB", "FANCC", "BRCA2", "FANCD2", "FANCE", "FANCF", "FANCG", "FANCI", "BRIP1", "FANCL", "FANCM", "PALB2", "RAD51C", "SLX4", "C1orf86", "C19orf40", "APITD1", "WDR48", "USP1", "C17orf70", "HES1", "RMI1", "RMI2", "TOP3A", "UBE2T", "XRCC6", "XRCC5", "PRKDC", "LIG4", "XRCC4", "DCLRE1C", "NHEJ1", "DNTT", "NUDT1", "DUT", "RRM2B", "POLB", "POLG", "POLD1", "POLE", "PCNA", "REV3L", "MAD2L2", "REV1", "POLH", "POLI", "POLQ", "POLK", "POLL", "POLM", "POLN", "FEN1", "FAN1", "TREX1", "TREX2", "EXO1", "APTX", "SPO11", "ENDOV", "UBE2A", "UBE2B", "RAD18", "SHPRH", "HLTF", "RNF168", "SPRTN", "RNF8", "RNF4", "UBE2V2", "UBE2N", "H2AFX", "CHAF1A", "SETMAR", "BLM", "WRN", "RECQL4", "ATM", "MPLKIP", "DCLRE1A", "DCLRE1B", "RPA4", "PRPF19", "RECQL", "RECQL5", "HELQ", "RDM1", "NABP2", "RFC1", "RFC2", "RFC3", "RFC4", "RFC5", "ATR", "ATRIP", "MDC1", "RAD1", "RAD9A", "HUS1", "RAD17", "CHEK1", "CHEK2", "TP53", "TP53BP1", "RIF1", "TOPBP1", "CLK2", "PER1")

# Filtering of variants based on DDR genes using SYMBOL 
exac_full_ddr.df <- fixed_af_filt_exac_full_ddr.df %>%
  filter(fixed_af_filt_exac_full_ddr.df$SYMBOL %in% ddr_genes)

exac_nontcga_ddr.df <- fixed_af_filt_exac_nontcga_ddr.df %>%
  filter(fixed_af_filt_exac_nontcga_ddr.df$SYMBOL %in% ddr_genes)

icr1000_ddr.df <- fixed_af_filt_icr1000_ddr.df %>%
  filter(fixed_af_filt_icr1000_ddr.df$SYMBOL %in% ddr_genes)

kgen_ddr.df <- fixed_af_filt_kgen_ddr.df %>%
  filter(fixed_af_filt_kgen_ddr.df$SYMBOL %in% ddr_genes)

# Check DDR gene filtered dataframes
dim(exac_full_ddr.df) # should be 88,952
dim(exac_nontcga_ddr.df) # should be 82,501
dim(icr1000_ddr.df) # should be 5,841
dim(kgen_ddr.df) # should be 271,929

# Clean-up
rm(fixed_af_filt_exac_full_ddr.df, fixed_af_filt_exac_nontcga_ddr.df, fixed_af_filt_icr1000_ddr.df, fixed_af_filt_kgen_ddr.df)

```

# transcript_variant_filtering

Filtering based on VEP consequences:  
1) synonymous_variant  
2) missense_variant  
3) inframe_insertion  
4) inframe_deletion  
5) stop_gained  
6) stop_lost  
7) stop_retained_variant  
8) start_lost  
9) frameshift_variant  
10) coding_sequence_variant  
11) protein_altering_variant  
12) 5_prime_UTR_variant  
13) 3_prime_UTR_variant

```{r transcript_variant_filtering}

# Filtration based on consequence
exac_full_tv_ddr.df <- exac_full_ddr.df %>% 
  filter(grepl("synonymous_variant",Consequence) |
           grepl("missense_variant",Consequence) |
           grepl("inframe_insertion", Consequence) |
           grepl("inframe_deletion", Consequence) |
           grepl("stop_gained",Consequence) |
           grepl("stop_lost",Consequence) |
           grepl("stop_retained_variant",Consequence) |
           grepl("start_lost",Consequence) |
           grepl("frameshift_variant",Consequence) |
           grepl("coding_sequence_variant",Consequence) |
           grepl("protein_altering_variant",Consequence) |
           grepl("5_prime_UTR_variant",Consequence) |
           grepl("3_prime_UTR_variant",Consequence))

exac_nontcga_tv_ddr.df <- exac_nontcga_ddr.df %>% 
  filter(grepl("synonymous_variant",Consequence) |
           grepl("missense_variant",Consequence) |
           grepl("inframe_insertion", Consequence) |
           grepl("inframe_deletion", Consequence) |
           grepl("stop_gained",Consequence) |
           grepl("stop_lost",Consequence) |
           grepl("stop_retained_variant",Consequence) |
           grepl("start_lost",Consequence) |
           grepl("frameshift_variant",Consequence) |
           grepl("coding_sequence_variant",Consequence) |
           grepl("protein_altering_variant",Consequence) |
           grepl("5_prime_UTR_variant",Consequence) |
           grepl("3_prime_UTR_variant",Consequence))

icr1000_tv_ddr.df <- icr1000_ddr.df %>% 
  filter(grepl("synonymous_variant",Consequence) |
           grepl("missense_variant",Consequence) |
           grepl("inframe_insertion", Consequence) |
           grepl("inframe_deletion", Consequence) |
           grepl("stop_gained",Consequence) |
           grepl("stop_lost",Consequence) |
           grepl("stop_retained_variant",Consequence) |
           grepl("start_lost",Consequence) |
           grepl("frameshift_variant",Consequence) |
           grepl("coding_sequence_variant",Consequence) |
           grepl("protein_altering_variant",Consequence) |
           grepl("5_prime_UTR_variant",Consequence) |
           grepl("3_prime_UTR_variant",Consequence))

kgen_tv_ddr.df <- kgen_ddr.df %>% 
  filter(grepl("synonymous_variant",Consequence) |
           grepl("missense_variant",Consequence) |
           grepl("inframe_insertion", Consequence) |
           grepl("inframe_deletion", Consequence) |
           grepl("stop_gained",Consequence) |
           grepl("stop_lost",Consequence) |
           grepl("stop_retained_variant",Consequence) |
           grepl("start_lost",Consequence) |
           grepl("frameshift_variant",Consequence) |
           grepl("coding_sequence_variant",Consequence) |
           grepl("protein_altering_variant",Consequence) |
           grepl("5_prime_UTR_variant",Consequence) |
           grepl("3_prime_UTR_variant",Consequence))

# Clean-up
rm(exac_full_ddr.df, exac_nontcga_ddr.df, icr1000_ddr.df, kgen_ddr.df, ddr_genes)

# Check output dataframes
unique(exac_full_tv_ddr.df$Consequence)
dim(exac_full_tv_ddr.df)
unique(exac_nontcga_tv_ddr.df$Consequence)
dim(exac_nontcga_tv_ddr.df)
unique(icr1000_tv_ddr.df$Consequence)
dim(icr1000_tv_ddr.df)
unique(kgen_tv_ddr.df$Consequence)
dim(kgen_tv_ddr.df)

```

# final_section

```{r final_section}

ls()
sessionInfo()
Sys.time()

```

